---
title: "Homework 2"
author: "Stephanie Chuquipul Calderon"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Problem 1

```{r import-subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

dim(subway_df)
```
This dataset contains information about NYC subway station entrances and exits, including the line, station name, station latitude and longitude, routes served, whether entry is allowed, vending availability, entrance type, and ADA compliance. I cleaned the column names, retained the required variables, and converted entry from "YES"/"NO" to TRUE/FALSE. The resulting dataset has `r nrow(subway_df)` rows and `r ncol(subway_df)` columns. These data are not yet fully tidy because routes served are spread across eleven columns rather than stored in a single route variable.

```{r count-stations}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```

```{r count-ada-stations}
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
Of the 465 distinct stations, 84 are ADA compliant.

```{r entry-without-vending}
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
37.7% of entrances/exits without vending allow entry.

```{r tidy-subway-routes}
subway_routes_df =
  subway_df |>
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_transform = list(route_name = as.character),
    values_drop_na = TRUE
  ) |>
  mutate(route_number = as.integer(route_number))

head(subway_routes_df)
```

```{r count-a-train-stations}
subway_routes_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

subway_routes_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
After reshaping the route columns into route number and route name, I identified 60 distinct stations serving the 
A train. Of these, 17 are ADA compliant.

## Problem 2
```{r clean_mr_trash_wheel}
mr_trash_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

nrow(mr_trash_df)
```

```{r clean_professor_trash_wheel}
professor_trash_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Professor Trash Wheel"
  )

nrow(professor_trash_df)
```

```{r clean_gwynnda}
gwynnda_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395",
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Gwynnda"
  )

nrow(gwynnda_df)
```

```{r combine_trash_wheels}
trash_wheel_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df)

nrow(trash_wheel_df)
```

```{r professor_total_weight}
professor_total_weight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

professor_total_weight
```

```{r gwynnda_june_cigarettes}
gwynnda_june_cigarettes =
  trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    month == "June",
    year == 2022
  ) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)

gwynnda_june_cigarettes
```

The combined dataset contains `r nrow(trash_wheel_df)` observations,
with each row representing a dumpster collected by Mr. Trash Wheel,
Professor Trash Wheel, or Gwynnda. Key variables include the collection
date, trash weight in tons, volume in cubic yards, and counts of items
such as plastic bottles and cigarette butts. I cleaned the column names,
removed rows without dumpster numbers, standardized year and month
values, and rounded Mr. Trash Wheel's sports ball counts to integers.
I added a variable identifying each wheel before combining the datasets.
Variables not recorded for a particular wheel are represented by missing
values. Professor Trash Wheel collected a total of
`r professor_total_weight` tons of trash, and Gwynnda collected
`r format(gwynnda_june_cigarettes, big.mark = ",")` cigarette butts
in June 2022.

## Problem 3



