---
title: "Homework 2"
author: "Stephanie Chuquipul Calderon"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Problem 1

```{r import-subway}
subway_df =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

dim(subway_df)
```
This dataset contains information about NYC subway station entrances and exits, including the line, station name, station latitude and longitude, routes served, whether entry is allowed, vending availability, entrance type, and ADA compliance. I cleaned the column names, retained the required variables, and converted entry from "YES"/"NO" to TRUE/FALSE. The resulting dataset has `r nrow(subway_df)` rows and `r ncol(subway_df)` columns. These data are not yet fully tidy because routes served are spread across eleven columns rather than stored in a single route variable.

```{r count-stations}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```

```{r count-ada-stations}
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
Of the 465 distinct stations, 84 are ADA compliant.

```{r entry-without-vending}
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
37.7% of entrances/exits without vending allow entry.

```{r tidy-subway-routes}
subway_routes_df =
  subway_df |>
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_transform = list(route_name = as.character),
    values_drop_na = TRUE
  ) |>
  mutate(route_number = as.integer(route_number))

head(subway_routes_df)
```

```{r count-a-train-stations}
subway_routes_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

subway_routes_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
After reshaping the route columns into route number and route name, I identified 60 distinct stations serving the 
A train. Of these, 17 are ADA compliant.

## Problem 2
```{r clean_mr_trash_wheel}
mr_trash_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N743"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

nrow(mr_trash_df)
```

```{r clean_professor_trash_wheel}
professor_trash_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M140",
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Professor Trash Wheel"
  )

nrow(professor_trash_df)
```

```{r clean_gwynnda}
gwynnda_df =
  readxl::read_excel(
    "data/trash_wheel.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395",
    na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Gwynnda"
  )

nrow(gwynnda_df)
```

```{r combine_trash_wheels}
trash_wheel_df =
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df)

nrow(trash_wheel_df)
```

```{r professor_total_weight}
professor_total_weight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

professor_total_weight
```

```{r gwynnda_june_cigarettes}
gwynnda_june_cigarettes =
  trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    month == "June",
    year == 2022
  ) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)

gwynnda_june_cigarettes
```

The combined dataset contains `r nrow(trash_wheel_df)` observations,
with each row representing a dumpster collected by Mr. Trash Wheel,
Professor Trash Wheel, or Gwynnda. Key variables include the collection
date, trash weight in tons, volume in cubic yards, and counts of items
such as plastic bottles and cigarette butts. I cleaned the column names,
removed rows without dumpster numbers, standardized year and month
values, and rounded Mr. Trash Wheel's sports ball counts to integers.
I added a variable identifying each wheel before combining the datasets.
Variables not recorded for a particular wheel are represented by missing
values. Professor Trash Wheel collected a total of
`r professor_total_weight` tons of trash, and Gwynnda collected
`r format(gwynnda_june_cigarettes, big.mark = ",")` cigarette butts
in June 2022.

## Problem 3

```{r clean_mci_baseline}
baseline_df =
  read_csv(
    "data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c("", "NA", ".")
  ) |>
  janitor::clean_names() |>
  rename(
    baseline_age = current_age,
    onset_age = age_at_onset
  ) |>
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  )

nrow(baseline_df)

baseline_df =
  baseline_df |>
  filter(is.na(onset_age) | onset_age > baseline_age)

nrow(baseline_df)
```

```{r baseline_summary}
baseline_df |>
  summarize(
    participants = n(),
    developed_mci = sum(!is.na(onset_age)),
    mean_baseline_age = mean(baseline_age, na.rm = TRUE)
  )

baseline_df |>
  filter(sex == "Female") |>
  summarize(
    proportion_apoe4 = mean(apoe4 == "Carrier")
  )
```

The baseline file contains 483 recruited participants. I skipped the
explanatory first row, cleaned the column names, and treated "." as a
missing value for MCI onset age. I encoded sex and APOE4 carrier status
as factors with descriptive labels. Four participants had MCI onset at
or before baseline and were excluded, leaving 479 eligible participants.
The cleaned dataset has one row per participant and includes baseline
age, sex, years of education, APOE4 status, and MCI onset age. Among the
eligible participants, 93 developed MCI during follow-up. The average
baseline age was 65.03 years, and 30% of women were APOE4 carriers.

```{r clean_amyloid}
amyloid_df =
  read_csv(
    "data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c("", "NA", "Na")
  ) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    cols = baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    time_years = case_match(
      visit,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    )
  ) |>
  select(id, time_years, amyloid_ratio)

dim(amyloid_df)
```

I imported the amyloid data by skipping the explanatory first row,
cleaning the column names, and treating both "NA" and "Na" as missing
values. I renamed the participant identifier to match the baseline
dataset. Using pivot_longer, I reshaped the five measurement columns
into separate variables for time since baseline and amyloid 42/40 ratio.
The tidy dataset contains 487 participants and 2,435 participant-time
rows, covering baseline and years 2, 4, 6, and 8. Missing biomarker
measurements are retained as NA.

```{r compare_participant_ids}
baseline_only_df =
  anti_join(baseline_df, amyloid_df, by = "id")

amyloid_only_df =
  amyloid_df |>
  distinct(id) |>
  anti_join(baseline_df, by = "id")

nrow(baseline_only_df)
nrow(amyloid_only_df)
```

After excluding participants with MCI at baseline, 8 participants in
the baseline dataset have no matching amyloid records, and 16 participants
in the amyloid dataset have no matching record in the eligible baseline
dataset.

```{r combine_mci_data}
mci_combined_df =
  inner_join(baseline_df, amyloid_df, by = "id")

dim(mci_combined_df)

mci_combined_df |>
  distinct(id) |>
  nrow()

write_csv(mci_combined_df, "data/mci_combined.csv")
```

I combined the eligible baseline and amyloid datasets using an inner
join on participant ID, retaining only participants present in both.
The resulting dataset contains 471 participants, with five scheduled
time points per participant, giving 2,355 rows and 8 columns. Each row
contains a participant's baseline demographics, MCI onset age, time
since baseline, and amyloid ratio. Missing biomarker measurements are
retained as NA. I exported the combined dataset to data/mci_combined.csv.



